---
title: "Study 1 (Heywood cases removed): Overall performance tables"
author: "Giacomo Bignardi"
cache: false
freeze: true
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    cache.lazy: false
---

# Background

This is a variant of `2_study1_2_analysis.R`'s two "Overall Performance" tables (test-retest reliability estimand, and population coefficient H estimand), re-run after removing Heywood cases for the EFA coefficient H estimator ("h"): any row with `est >= 1` for that estimator is dropped before summarising.

Analyses are run inside the [`bignardig/tidyverse461:v5`](https://hub.docker.com/r/bignardig/tidyverse461) Docker container.

```{r setup}
library(tidyverse)
library(gt)

rm(list = ls(all.names = TRUE))
```

# Load Data

## Loading conditions used in the simulation

```{r loading-conditions}
loadings_list = list(
  c( 0, 0, 0, 0, 0, 0),
  c(.1,.1,.1,.1,.1),
  c(.3,.2,.1),
  c(.4,.3,.3,.2,.1,.0),
  c(.4,.4,.4,.4),
  c(.6,.5,.4,.3,.2),
  c(.7,.6,.5,.5,.5,.4,.4,.3,.3)
)

loadings_list_paste = lapply(loadings_list, function(x) paste0(x, collapse = "_")) %>% unlist()

# All-zero loadings condition: no true common factor, so the sampled
# test-retest correlation is pure noise around 0. Indexed by loading_set,
# used below to clamp the estimand to 0.001 in that condition (matching
# 2_study1_2_analysis.R) instead of a noisy sample mean.
loadings_all_zero = sapply(loadings_list, function(x) all(x == 0))
```

## Read raw simulation results

```{r load-results}
# I ran the simulation code multiple times on the cluster, so we have several results files we want to join

results_path = file.path("results","study1_results")
results_files = list.files(results_path,
                           pattern = ".rds",
                           recursive = FALSE,
                           full.names = TRUE
                           )
results = lapply(results_files, function(x) readRDS(x))
results = do.call("c", results)
```

## Build results_table (wide, one row per replicate)

```{r build-results-table}
results_table = data.frame(i = 1:length(results))

results_table$sample_sizes    = sapply(results, function(x) x$settings$n) %>% as.numeric()
results_table$n_items         = sapply(results, function(x) length(x$settings$loadings)) %>% as.numeric()
results_table$loadings        = sapply(results, function(x) paste0(x$settings$loadings, collapse = "_"))
results_table$loading_set     = match(results_table$loadings , loadings_list_paste)

results_table$pop_coefh         = sapply(results, function(x) x$population_reliability) %>% as.numeric()    # Population coefficient H
results_table$pop_ss_loading    = sapply(results, function(x) sum(x$settings$loadings^2)) %>% as.numeric()

# Test-retest reliability: correlation between the Bayesian factor scores estimated
# at t1 and t2.
results_table$test_retest_reliability = sapply(results, function(x) x$test_retest_reliability) %>% as.numeric()

### T1 estimates
results_table$rmu_est_t1  = sapply(results, function(x) x$rmu_est_t1$rmu_estimate) %>% as.numeric()
results_table$rmu_lb_t1   = sapply(results, function(x) x$rmu_est_t1$hdci_lowerbound) %>% as.numeric()
results_table$rmu_ub_t1   = sapply(results, function(x) x$rmu_est_t1$hdci_upperbound) %>% as.numeric()

results_table$h_est_t1    = sapply(results, function(x) x$h_reliability_t1$r) %>% as.numeric()
results_table$h_lb_t1     = sapply(results, function(x) x$h_reliability_t1$ci[1]) %>% as.numeric()
results_table$h_ub_t1     = sapply(results, function(x) x$h_reliability_t1$ci[2]) %>% as.numeric()

# a_lb_t1/a_ub_t1 recomputed from est +/- 1.96*se rather than MBESS's own
# $ci.lower/$ci.upper, since MBESS::ci.reliability() clamps its reported
# bounds to [0, 1] without adjusting $est - see 2_study1_2_analysis.R.
results_table$a_est_t1    = sapply(results, function(x) x$alpha_reliability_t1$est) %>% as.numeric()
results_table$a_se_t1     = sapply(results, function(x) x$alpha_reliability_t1$se)  %>% as.numeric()
results_table$a_lb_t1     = results_table$a_est_t1 - qnorm(0.975)*results_table$a_se_t1
results_table$a_ub_t1     = results_table$a_est_t1 + qnorm(0.975)*results_table$a_se_t1

results_table$irth_est_t1 = sapply(results, function(x) x$mcmc_coefh_t1$mcmc_coef_h) %>% as.numeric()
results_table$irth_lb_t1  = sapply(results, function(x) x$mcmc_coefh_t1$.lower) %>% as.numeric()
results_table$irth_ub_t1  = sapply(results, function(x) x$mcmc_coefh_t1$.upper) %>% as.numeric()

results_table$true_score_coverage_t1 = sapply(results, function(x) x$true_score_coverage_t1) %>% as.numeric()
results_table$diag_divergences_t1    = sapply(results, function(x) x$diagnostics_divergences_t1) %>% as.numeric()
results_table$diag_ebfmi_t1          = sapply(results, function(x) length(which(x$diagnostics_ebfmi_t1 < .2)))

### T2 estimates
results_table$rmu_est_t2  = sapply(results, function(x) x$rmu_est_t2$rmu_estimate) %>% as.numeric()
results_table$rmu_lb_t2   = sapply(results, function(x) x$rmu_est_t2$hdci_lowerbound) %>% as.numeric()
results_table$rmu_ub_t2   = sapply(results, function(x) x$rmu_est_t2$hdci_upperbound) %>% as.numeric()

results_table$h_est_t2    = sapply(results, function(x) x$h_reliability_t2$r) %>% as.numeric()
results_table$h_lb_t2     = sapply(results, function(x) x$h_reliability_t2$ci[1]) %>% as.numeric()
results_table$h_ub_t2     = sapply(results, function(x) x$h_reliability_t2$ci[2]) %>% as.numeric()

results_table$a_est_t2    = sapply(results, function(x) x$alpha_reliability_t2$est) %>% as.numeric()
results_table$a_se_t2     = sapply(results, function(x) x$alpha_reliability_t2$se)  %>% as.numeric()
results_table$a_lb_t2     = results_table$a_est_t2 - qnorm(0.975)*results_table$a_se_t2
results_table$a_ub_t2     = results_table$a_est_t2 + qnorm(0.975)*results_table$a_se_t2

results_table$irth_est_t2 = sapply(results, function(x) x$mcmc_coefh_t2$mcmc_coef_h) %>% as.numeric()
results_table$irth_lb_t2  = sapply(results, function(x) x$mcmc_coefh_t2$.lower) %>% as.numeric()
results_table$irth_ub_t2  = sapply(results, function(x) x$mcmc_coefh_t2$.upper) %>% as.numeric()

results_table$true_score_coverage_t2 = sapply(results, function(x) x$true_score_coverage_t2) %>% as.numeric()
results_table$diag_divergences_t2    = sapply(results, function(x) x$diagnostics_divergences_t2) %>% as.numeric()
results_table$diag_ebfmi_t2          = sapply(results, function(x) length(which(x$diagnostics_ebfmi_t2 < .2)))

results_table$sample_sizes = factor(results_table$sample_sizes)
```

# Reshape to Long Format

```{r reshape-long}
## Stack t1/t2 into long format
# `rowid` identifies the original replicate and is shared across its t1/t2 rows.

results_table_stacked = results_table %>%
  rename(rowid = i) %>%
  pivot_longer(
    cols = matches("_t[12]$"),
    names_to = c(".value", "time"),
    names_pattern = "(.*)_t([12])"
  ) %>%
  mutate(
    time = paste0("t", time),
    diag_divergences_binary = as.numeric(diag_divergences > 0),
    diag_ebfmi_binary       = as.numeric(diag_ebfmi > 0)
  )

## Pivot longer over estimator name
results_table_long = results_table_stacked %>%
  pivot_longer(cols = c(rmu_est, h_est, a_est, irth_est,
                        rmu_lb, h_lb, a_lb, irth_lb,
                        rmu_ub, h_ub, a_ub, irth_ub
                        ), names_to = c("name", ".value"), names_pattern = "(rmu|h|a|irth)_(.*)")
```

# Remove Heywood Cases (EFA Coefficient H)

Heywood cases for the factor analysis H estimator ("h") show up as reliability estimates `>= 1`. We drop those rows before summarising, leaving the other three estimators (RMU, coefficient alpha, IRT H) untouched.

```{r remove-heywood}
n_heywood = results_table_long %>%
  filter(name == "h", est >= 1) %>%
  nrow()

results_table_long = results_table_long %>%
  filter(!(name == "h" & est >= 1))

n_heywood
```

`r n_heywood` Heywood case(s) (EFA coefficient H estimates `>= 1`) were removed.

# Performance Metrics

## Overall Performance: test-retest reliability estimand

```{r overall-performance-testretest}
results_table_long %>%
  group_by(name, loading_set, sample_sizes) %>%
  mutate(
    name       = factor(name, levels = c("rmu","a","h","irth"),
                        labels = c("RMU", "coefficient alpha", "factor analysis H", "IRT H")),
    estimand   = ifelse(loadings_all_zero[loading_set], 0.001, mean(test_retest_reliability))        # mean test-retest correlation within this sample size / loading condition; clamped to 0.001 when loadings are all zero (as in 2_study1_2_analysis.R)
  ) %>%
  ungroup() %>%
  group_by(name) %>%
  mutate(
    difference = est - estimand,
    ci_correct = (lb <= estimand & ub >= estimand),
    ci_length  = ub - lb
  ) %>%
  summarise(

    n              = n(),

    # Empirical Standard Error (point estimate only)
    EmpSE          = sd(est),

    # Bias
    bias           = mean(difference),

    # Root Mean Squared Error
    RMSE           = sqrt(mean((difference)^2)),

    # Coverage
    coverage       = length(which(ci_correct))/length(ci_correct),

    mean_ci_length = mean(ci_length)

  ) %>%
  ungroup() %>%
  select(name, n, RMSE, bias, EmpSE, coverage, mean_ci_length) %>%
  gt() %>%
  gt::fmt(columns = !c(name, n), fns = ~ gbtoolbox::apa_num(., n_decimal_places = 3)) %>%
  cols_label(
    mean_ci_length ~ "Avg CI Length"
  ) %>%
  tab_footnote(
    footnote = html("<b>n</b> = number of simulations included after removing Heywood cases (t1 and t2 estimates counted separately).
                <b>RMSE</b> = Root Mean Squared Error of the estimate around the estimand.
                <b>bias</b> = mean(estimate &minus; estimand).
                <b>EmpSE</b> = empirical standard deviation of the estimates across simulations.
                <b>coverage</b> = proportion of simulations where the estimator's own 95% CI/CrI contained the estimand, which should be close to 95%.
                <b>Avg CI Length</b> = mean width of the estimator's own 95% CI/CrI.
                <b>estimand</b> = test-retest reliability, i.e. the correlation between the Bayesian factor scores estimated at t1 and t2, averaged within each sample size / loading condition."
    )
  )
```

## Maximal Reliability: population coefficient H estimand

```{r overall-performance-coefh}
results_table_long %>%
  mutate(
    name       = factor(name, levels = c("rmu","a","h","irth"),
                        labels = c("RMU", "coefficient alpha", "factor analysis H", "IRT H")),
    estimand   = ifelse(pop_coefh == 0, 0.001, pop_coefh),      # clamped to 0.001 when population coefficient H is zero (as in 2_study1_2_analysis.R)
    difference = est - estimand,
    ci_correct = (lb <= estimand & ub >= estimand),
    ci_length  = ub - lb,
  ) %>%
  group_by(name) %>%
  summarise(

    n              = n(),

    # Empirical Standard Error (point estimate only)
    EmpSE          = sd(est),

    # Bias
    bias           = mean(difference),

    # Root Mean Squared Error
    RMSE           = sqrt(mean((difference)^2)),

    # Coverage
    coverage       = length(which(ci_correct))/length(ci_correct),

    mean_ci_length = mean(ci_length)

  ) %>%
  ungroup() %>%
  select(name, n, RMSE, bias, EmpSE, coverage, mean_ci_length) %>%
  gt() %>%
  gt::fmt(columns = !c(name, n), fns = ~ gbtoolbox::apa_num(., n_decimal_places = 3)) %>%
  cols_label(
    mean_ci_length ~ "Avg CI Length"
  ) %>%
  tab_footnote(
    footnote = html("<b>n</b> = number of simulations included after removing Heywood cases (t1 and t2 estimates counted separately).
                <b>RMSE</b> = Root Mean Squared Error of the estimate around the estimand.
                <b>bias</b> = mean(estimate &minus; estimand).
                <b>EmpSE</b> = empirical standard deviation of the estimates across simulations.
                <b>coverage</b> = proportion of simulations where the estimator's own 95% CI/CrI contained the estimand, which should be close to 95%.
                <b>Avg CI Length</b> = mean width of the estimator's own 95% CI/CrI.
                <b>estimand</b> = population coefficient H (maximal reliability)."
    )
  )
```
