---
title: "Applied Example: Estimating reliability of reinforcement learning parameters using data from Pike et al 2026"
author: "Giacomo Bignardi"
cache: false
freeze: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

Claude Code was used to prepare the first draft of this document, which was later checked and edited. 

# Background

All data from this script comes from [Pike, A. C., … Roiser, J. P. (2026). The recoverability, reliability, and generalizability of reward processing parameters and relation to mental health symptoms. Psychological Medicine, 56, e238. doi:10.1017/S0033291726105340](https://www.cambridge.org/core/journals/psychological-medicine/article/recoverability-reliability-and-generalizability-of-reward-processing-parameters-and-relation-to-mental-health-symptoms/C909E875D2E161CE597FCDCEBD704562#article).

The [models](https://osf.io/p648q/overview) and [datasets](https://osf.io/kde4b/files/osfstorage) are available on the open science framework. To download the models, you may need to press the "Models" tab in Files Preview. 

We use data from the Fluctuating Band task (task A). Rather than fitting the model used in the paper, I've selected a simpler model (model 3) reported in the supplement and I've made some slight edits to the stan code to improve convergence. The stan code is printed in full in the [appendix below](#appendix-the-stan-model), alongside a walkthrough of how it works.

# Setup

Analyses are run inside the [`bignardig/tidyverse461:v5`](https://hub.docker.com/r/bignardig/tidyverse461) Docker container - which includes all dependencies except the cmdstan install (v2.39.0). 

The script `02_01_pike_2026_fit_bandit_models.R` fits the main Bayesian models to time 1 and time 2 separately. It models only the N = 115 subjects who completed **both** sessions (the `bandit_t1.json` / `bandit_t2.json` sub sample, not the full N = 547 `bandit_all.json` sample).

The script `02_04_pike_2026_fit_bandit_splithalf_models.R` fits the Bayesian models four times, splitting the data each each timepoint into halves to estimate split-half reliability at each time point. It models only hte N = 115 subjects who completed **both** sessions. 

The script `02_01_pike_2026_fit_bandit_fullsample.R` fits the Bayesian model to the full (N = 547) dataset and also to split-half datasets to estimate RMU and split-half reliability on the full sample. 


```{r setup}
rm(list = ls())
library(gbtoolbox)
library(cmdstanr)
library(bayesplot)
library(tidyverse)
library(gt)

data_dir = file.path("data", "pike_2026")
subj_pars = c("lrR", "lrP", "ivT")
```

# Data

A look at the raw data files behind this notebook

## Raw trial-level exports

`bandit.csv` / `bandit2.csv` are the authors' raw per-trial exports for session 1 / session 2 (one row per trial, all subjects -- not yet subset to the N = 115 retest sample used elsewhere in this notebook). I've filtered down to the columns actually used for modelling.

```{r load-raw-bandit-data}
#| output: false

setwd("..") # set the working directory to the parent folder which has /data/

bandit_model_cols = c("id", "trial_nr", "is_gain", "is_loss", "response")

bandit_raw_t1 = read_csv(file.path(data_dir, "bandit.csv"), show_col_types = FALSE) %>%
  select(all_of(bandit_model_cols))
bandit_raw_t2 = read_csv(file.path(data_dir, "bandit2.csv"), show_col_types = FALSE) %>%
  select(all_of(bandit_model_cols))

```

Time 1

```{r table-raw-bandit-t1}
rmarkdown::paged_table(bandit_raw_t1, options = list(rows.print = 25, cols.print = ncol(bandit_raw_t1)))
```

Time 2

```{r table-raw-bandit-t2}
rmarkdown::paged_table(bandit_raw_t2, options = list(rows.print = 25, cols.print = ncol(bandit_raw_t2)))
```

## Stan-ready data lists

`bandit_t1.json` / `bandit_t2.json` are the modeling-ready data lists for the matched N = 115 subjects that `02_01_pike_2026_fit_bandit_models.R` passes to `model_3_noncentered.stan` -- built by the authors' `bandit_datalist()` (`data/pike_2026/utils.R`) from the raw csvs above.

```{r load-bandit-json}
library(jsonlite)

setwd("..") # set the working directory to the parent folder which has /data/

bandit_json_t1 = fromJSON(file.path(data_dir, "bandit_t1.json"))
bandit_json_t2 = fromJSON(file.path(data_dir, "bandit_t2.json"))

str(bandit_json_t1, list.len = 10)
```

```{r inspect-bandit-json}
#| include: false

# first subject, first 10 trials: choice (1-4) and the reward/punishment received
bandit_json_t1$choice[1, 1:10]
bandit_json_t1$rwd[1, 1:10]
bandit_json_t1$plt[1, 1:10]
```
## Load saved fits

```{r load-saved-fits}

setwd("..") # set the working directory to the parent folder which has /data/

fits_dir = file.path(data_dir, "model_fits")

if (!file.exists(file.path(fits_dir, "fit_model3nc_bandit_t1.RDS"))) {
  stop("No saved fits found. Run 02_01_pike_2026_fit_bandit_models.R first.")
}
if (!file.exists(file.path(fits_dir, "fit_model3nc_bandit_fullsample.RDS"))) {
  stop("No full-sample fit found. Run 02_01_pike_2026_fit_bandit_fullsample.R first.")
}
if (!file.exists(file.path(fits_dir, "fit_model3nc_bandit_t1_first.RDS"))) {
  stop("No split-half fits found. Run 02_04_pike_2026_fit_bandit_splithalf_models.R first.")
}
if (!file.exists(file.path(fits_dir, "fit_model3nc_bandit_fullsample_first.RDS"))) {
  stop("No full-sample split-half fits found. Run 02_01_pike_2026_fit_bandit_fullsample.R first.")
}

# retest sample (N = 115), time 1 / time 2
ids_t1 = readRDS(file.path(fits_dir, "bandit_ids_t1t2.RDS"))
ids_t2 = ids_t1 # same 115 subjects, same order -- enforced when the JSONs were built

fit_bandit_t1 = readRDS(file.path(fits_dir, "fit_model3nc_bandit_t1.RDS"))
fit_bandit_t2 = readRDS(file.path(fits_dir, "fit_model3nc_bandit_t2.RDS"))

# full T1 sample (N = 547)
ids_all = readRDS(file.path(fits_dir, "bandit_ids_all.RDS"))
fit_bandit_all = readRDS(file.path(fits_dir, "fit_model3nc_bandit_fullsample.RDS"))

# retest sample split-half (N = 115), time 1 / time 2
ids_splithalf = readRDS(file.path(fits_dir, "bandit_splithalf_ids.RDS"))

fit_bandit_t1_first  = readRDS(file.path(fits_dir, "fit_model3nc_bandit_t1_first.RDS"))
fit_bandit_t1_second = readRDS(file.path(fits_dir, "fit_model3nc_bandit_t1_second.RDS"))
fit_bandit_t2_first  = readRDS(file.path(fits_dir, "fit_model3nc_bandit_t2_first.RDS"))
fit_bandit_t2_second = readRDS(file.path(fits_dir, "fit_model3nc_bandit_t2_second.RDS"))

# full T1 sample split-half (N = 547)
fit_bandit_fullsample_first  = readRDS(file.path(fits_dir, "fit_model3nc_bandit_fullsample_first.RDS"))
fit_bandit_fullsample_second = readRDS(file.path(fits_dir, "fit_model3nc_bandit_fullsample_second.RDS"))

```

# Diagnostics

We fit `model_3_noncentered.stan` nine times across this notebook -- the T1/T2 retest-sample fits, the full T1 sample, and split-half fits for each of those three -- so before trusting any of the reliability estimates below, we check convergence across all nine, not just the two headline T1/T2 fits.

```{r all-fits}
all_fits = list(
  "T1 (retest, N = 115)"          = fit_bandit_t1,
  "T2 (retest, N = 115)"          = fit_bandit_t2,
  "Full T1 sample (N = 547)"      = fit_bandit_all,
  "T1 first-half (N = 115)"       = fit_bandit_t1_first,
  "T1 second-half (N = 115)"      = fit_bandit_t1_second,
  "T2 first-half (N = 115)"       = fit_bandit_t2_first,
  "T2 second-half (N = 115)"      = fit_bandit_t2_second,
  "Full T1 first-half (N = 547)"  = fit_bandit_fullsample_first,
  "Full T1 second-half (N = 547)" = fit_bandit_fullsample_second
)

hyper_pars = c("mu", "sigma")
diag_pars  = c(hyper_pars, subj_pars)

rhat_threshold = 1.01
ess_threshold  = 400 # cmdstanr's own rule of thumb: bulk/tail ESS < 100 per chain (400 across our 4 chains)
```

## Sampler diagnostics (divergences, treedepth, E-BFMI)

```{r sampler-diagnostics}
sampler_diag = imap_dfr(all_fits, function(fit, label) {
  d = fit$diagnostic_summary()
  tibble(
    model           = label,
    n_divergent     = sum(d$num_divergent),
    n_max_treedepth = sum(d$num_max_treedepth),
    min_ebfmi       = min(d$ebfmi)
  )
})

sampler_diag %>%
  gt() %>%
  fmt_number(columns = min_ebfmi, decimals = 2) %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = n_divergent, rows = n_divergent > 0)
  ) %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = n_max_treedepth, rows = n_max_treedepth > 0)
  ) %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = min_ebfmi, rows = min_ebfmi < 0.3)
  )
```

All nine fits show zero divergences, zero max-treedepth hits, and E-BFMI comfortably above the 0.3 warning threshold -- any red cell above would mean one of the nine fits needs a closer look before its reliability numbers are trusted.

## Rhat / ESS by parameter, across all models

Rather than eyeballing `summary()` output fit-by-fit, we collapse each fit's per-subject `rhat`/`ess_bulk`/`ess_tail` down to one row per parameter family (`mu`, `sigma`, `lrR`, `lrP`, `ivT`) -- taking the worst (max rhat / min ESS) across subjects within that family -- so a single bad subject-level parameter can't hide inside a fit-wide average. Flagged cells breach the conventional rhat > 1.01 / ESS < 400 thresholds.

```{r param-diagnostics}
param_diag = imap_dfr(all_fits, function(fit, label) {
  fit$summary(variables = diag_pars) %>%
    mutate(parameter = str_extract(variable, "^[^\\[]+")) %>%
    group_by(parameter) %>%
    summarise(
      max_rhat     = max(rhat, na.rm = TRUE),
      min_ess_bulk = min(ess_bulk, na.rm = TRUE),
      min_ess_tail = min(ess_tail, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(model = label, .before = 1)
}) %>%
  mutate(parameter = factor(parameter, levels = diag_pars))

param_diag %>%
  gt(groupname_col = "model") %>%
  fmt_number(columns = max_rhat, decimals = 3) %>%
  fmt_number(columns = c(min_ess_bulk, min_ess_tail), decimals = 0) %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = max_rhat, rows = max_rhat > rhat_threshold)
  ) %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = min_ess_bulk, rows = min_ess_bulk < ess_threshold)
  ) %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = min_ess_tail, rows = min_ess_tail < ess_threshold)
  )
```

## Parameters with consistently poor convergence

Collapsing the table above across models: for each parameter family, the worst rhat/ESS seen in *any* of the nine fits, which fit produced it, and how many fits flag it -- so a parameter that's borderline everywhere is distinguishable from one bad fit dragging down an otherwise clean parameter.

```{r param-diagnostics-summary}
param_summary = param_diag %>%
  group_by(parameter) %>%
  summarise(
    worst_rhat            = max(max_rhat),
    worst_rhat_model       = model[which.max(max_rhat)],
    worst_ess_bulk         = min(min_ess_bulk),
    worst_ess_bulk_model   = model[which.min(min_ess_bulk)],
    n_models_rhat_flagged = sum(max_rhat > rhat_threshold),
    n_models_ess_flagged  = sum(min_ess_bulk < ess_threshold),
    .groups = "drop"
  ) %>%
  arrange(desc(worst_rhat))

param_summary %>%
  gt() %>%
  fmt_number(columns = worst_rhat, decimals = 3) %>%
  fmt_number(columns = worst_ess_bulk, decimals = 0) %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = n_models_rhat_flagged, rows = n_models_rhat_flagged > 0)
  ) %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = n_models_ess_flagged, rows = n_models_ess_flagged > 0)
  )
```

## Which subjects are driving the poor rhat/ESS?

The per-parameter table above takes the *worst* subject in each fit -- it can't tell us whether the same handful of subjects are consistently unidentifiable (worth flagging, or excluding, individually) or whether the problem is spread thinly across many different subjects (more likely a sampler-tuning issue than a subject-data issue). We rebuild the same rhat/ESS check but keep every subject, then rank subjects by how often they're flagged across all the (fit x parameter) combinations they appear in.

```{r subject-level-diagnostics}
all_ids = list(
  "T1 (retest, N = 115)"          = ids_t1,
  "T2 (retest, N = 115)"          = ids_t2,
  "Full T1 sample (N = 547)"      = ids_all,
  "T1 first-half (N = 115)"       = ids_splithalf,
  "T1 second-half (N = 115)"      = ids_splithalf,
  "T2 first-half (N = 115)"       = ids_splithalf,
  "T2 second-half (N = 115)"      = ids_splithalf,
  "Full T1 first-half (N = 547)"  = ids_all,
  "Full T1 second-half (N = 547)" = ids_all
)

subject_diag = imap_dfr(all_fits, function(fit, label) {
  ids = all_ids[[label]]
  fit$summary(variables = subj_pars) %>%
    mutate(
      parameter = str_extract(variable, "^[^\\[]+"),
      idx       = as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")),
      id        = ids[idx]
    ) %>%
    select(id, parameter, rhat, ess_bulk, ess_tail) %>%
    mutate(model = label, .before = 1)
}) %>%
  mutate(
    parameter = factor(parameter, levels = subj_pars),
    flagged   = rhat > rhat_threshold | ess_bulk < ess_threshold
  )

subject_flag_summary = subject_diag %>%
  group_by(id) %>%
  summarise(
    n_evaluated    = n(), # how many (fit x parameter) combinations this subject appears in
    n_flagged      = sum(flagged),
    pct_flagged    = n_flagged / n_evaluated,
    worst_rhat     = max(rhat),
    worst_ess_bulk = min(ess_bulk),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_flagged), desc(worst_rhat))

subject_flag_summary %>%
  slice_head(n = 20) %>%
  gt() %>%
  fmt_percent(columns = pct_flagged, decimals = 0) %>%
  fmt_number(columns = worst_rhat, decimals = 3) %>%
  fmt_number(columns = worst_ess_bulk, decimals = 0) %>%
  tab_header(title = "20 subjects flagged most often (rhat > 1.01 or ESS < 400)") %>%
  tab_style(
    style     = list(cell_fill(color = "#f8d7da"), cell_text(weight = "bold")),
    locations = cells_body(columns = pct_flagged, rows = pct_flagged > 0.5)
  )
```

If the same subject IDs sit at the top of this table across most rows, that points to specific subjects (e.g. near-random responders where `ivT`/learning rates are only weakly identified by the likelihood) rather than a broad sampler problem. The histogram below shows whether flagged subjects are a small concentrated group or spread across the sample:

```{r subject-diagnostics-distribution}
subject_flag_summary %>%
  ggplot(aes(x = pct_flagged)) +
  geom_histogram(binwidth = 0.1, boundary = 0, fill = "steelblue", color = "white") +
  labs(
    x = "Proportion of (fit x parameter) combinations flagged",
    y = "Number of subjects",
    title = "How concentrated is the convergence problem?",
    subtitle = paste0(
      sum(subject_flag_summary$pct_flagged > 0.5), " of ", nrow(subject_flag_summary),
      " subjects are flagged in more than half of the fits/parameters they appear in"
    )
  ) +
  theme_minimal()
```

## Group-level parameter trace plots (T1 / T2 retest fits)

`mu`/`sigma` are the population mean/SD (probit scale) behind `lrR`/`lrP`/`ivT`, in that order -- i.e. `mu[2]`/`sigma[2]` are what determine where `lrP` gets pulled to and how tightly (see the per-subject posterior density discussion below).

```{r trace-plots-hyperparameters}
mcmc_trace(fit_bandit_t1$draws(variables = hyper_pars))
mcmc_trace(fit_bandit_t2$draws(variables = hyper_pars))
```
# Inspect subject-level parameter estimates

## Posterior Means

In this section, we extract posterior mean estimates for each subject's reward learning rate (`lrR`), punishment learning rate (`lrP`) and inverse-temperature (`ivT`).

```{r subject-level-estimates}
extract_indpars = function(fit, ids, pars = subj_pars) {
  fit$summary(variables = pars) %>% # This gives a long dataframe with the poster mean/median/quantiles for each parameter in the model
    mutate(
      parameter = str_extract(variable, "^[^\\[]+"),
      idx       = as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")),
      id        = ids[idx]
    ) %>%
    select(id, parameter, mean, sd) %>%
    pivot_wider(id_cols = id, names_from = parameter, values_from = mean)
}

indpars_t1 = extract_indpars(fit_bandit_t1, ids_t1)
indpars_t2 = extract_indpars(fit_bandit_t2, ids_t2)

indpars_t1
```

## Posterior densities

Below we show the posterior densities for each subject, separately for each time point. 

```{r per-subject-densities}
get_draws_long = function(fit, ids, pars = subj_pars) {
  fit$draws(variables = pars, format = "df") %>%
    as_tibble() %>%
    select(-.chain, -.iteration, -.draw) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
    mutate(
      parameter = str_extract(variable, "^[^\\[]+"),
      idx       = as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")),
      id        = ids[idx]
    )
}

plot_subject_densities = function(fit, ids, session_label, pars = subj_pars, ymax_cap = 20) {
  # browser()
  draws_long = get_draws_long(fit, ids, pars)

  # per-facet y upper limit: each subject's own density peak, capped at ymax_cap
  y_scales = draws_long %>%
    group_by(parameter, id) %>%
    summarise(peak = max(density(value)$y), .groups = "drop") %>%
    group_by(parameter) %>%
    summarise(ymax = min(max(peak), ymax_cap), .groups = "drop") %>%
    arrange(parameter) %>%
    pull(ymax) %>%
    purrr::map(~ scale_y_continuous(limits = c(0, .x)))

  draws_long %>%
    ggplot(aes(
      x = value,
      # y = after_stat(scaled),
      group = id)
      ) +
    geom_density(alpha = 0.1, linewidth = 0.3, color = "steelblue") +
    facet_wrap(~parameter, scales = "free") +
    ggh4x::facetted_pos_scales(y = y_scales) +
    labs(
      title = paste0("Per-subject posterior densities — ", session_label),
      subtitle = "One density curve per subject, each scaled to its own peak",
      x = NULL, y = "Density (scaled to subject's peak)"
    ) +
    theme_minimal()
}

plot_subject_densities(fit_bandit_t1, ids_t1, "T1 (N = ")
plot_subject_densities(fit_bandit_t2, ids_t2, "T2")
```

# Test-retest reliability

Correlate each subject's posterior-mean parameter estimate across the two sessions.

```{r empirical-test-retest}
stopifnot(identical(indpars_t1$id, indpars_t2$id))

testretest = map_dfr(subj_pars, function(p) {
  ct = cor.test(indpars_t1[[p]], indpars_t2[[p]])
  r  = unname(ct$estimate)
  tibble(
    parameter = p,
    method    = "Test-Retest (N = 115)",
    estimate  = r,
    se        = sqrt((1 - r^2) / ct$parameter), # SE of Pearson r implied by the t-test df (matches cor.test's statistic)
    lower     = ct$conf.int[1],
    upper     = ct$conf.int[2],
    width     = ct$conf.int[2] - ct$conf.int[1]
  )
})

testretest |> 
  gt() |> 
  fmt_number(decimals = 3)
  
```

# RMU

`gbtoolbox::reliability()` estimates a *within-session* reliability (`rmu`) using the posterior samples from each subject. 

## Retest sample (N = 115)

```{r posterior-reliability}
get_draws_matrix = function(fit, ids, par) {
  m = fit$draws(variables = par, format = "matrix") # draws x N
  colnames(m) = ids
  t(m) # N x draws
}

extract_rmu = function(fit, ids, session_label, pars = subj_pars) {
  map_dfr(pars, function(p) {
    # browser()
    rel = reliability(get_draws_matrix(fit, ids, p))
    tibble(
      parameter = p,
      method    = paste0("RMU (", session_label, ")"),
      estimate  = rel$hdci$rmu_estimate,
      se        = sd(rel$reliability_posterior_draws), # posterior SD of the RMU draws
      lower     = rel$hdci$hdci_lowerbound,
      upper     = rel$hdci$hdci_upperbound,
      width     = rel$hdci$hdci_upperbound - rel$hdci$hdci_lowerbound
    )
  })
}

rmu = bind_rows(
  extract_rmu(fit_bandit_t1, ids_t1, "T1, N = 115"),
  extract_rmu(fit_bandit_t2, ids_t2, "T2, N = 115")
)

rmu |> 
  gt() |> 
  fmt_number(decimals = 3)
```

To calculate RMU more simply (if you just want to check a single parameter), you need to extract a matrix of posterior draws with participants on the rows and samples on the columns, then pass to `gbtoolbox::reliability()`.

```{r}

results = fit_bandit_t1$draws(variables = "lrR", format = "matrix") |>  #this is our fitted stan model
  t() |> # make sure participants are on rows and that draws are on columns 
  gbtoolbox::reliability()

results

```

## Full T1 sample (N = 547)

RMU only needs one session's posterior draws -- `gbtoolbox::reliability()` never touches session 2 -- so it isn't limited to the N = 115 test-retest subsample used above. `bandit_all.json` is the full session-1 sample (N = 547, built in the authors' `Datalists.Rmd` as `bandit_datalist(bandit %>% arrange(id))`, unfiltered by session-2 completion), so fitting the same model to it gives an RMU estimate at the largest N the raw data actually supports. Run `02_01_pike_2026_fit_bandit_fullsample.R` first to produce this fit -- it isn't part of the test-retest/split-half fits above, so it has to be fit separately.

```{r load-full-sample-fit}
fit_bandit_all$diagnostic_summary()
```

```{r full-sample-rmu}
rel_full_sample = extract_rmu(fit_bandit_all, ids_all, paste0("Full T1 sample, N = ", length(ids_all)))

rel_full_sample |>
  gt() |>
  fmt_number(decimals = 3)
```

# Split-half reliability

In this section, we splitted the data from each session into two equally sized halves. We then fit the models on each split-half dataset independently. 

The correlation between each split-half dataset estimates the reliability of each split-half dataset, not of the combined data. To estimate the reliability of the full length test, we use the Spearman-Brown formula. 

## Retest sample (N = 115)

```{r empirical-splithalf}
spearman_brown = function(r) 2 * r / (1 + r)

splithalf_session = function(fit_first, fit_second, ids, session_label) {
  indpars_first  = extract_indpars(fit_first,  ids)
  indpars_second = extract_indpars(fit_second, ids)
  stopifnot(identical(indpars_first$id, indpars_second$id))

  map_dfr(subj_pars, function(p) {
    ct   = cor.test(indpars_first[[p]], indpars_second[[p]])
    r    = unname(ct$estimate)
    se_r = sqrt((1 - r^2) / ct$parameter) # SE of the underlying split-half r
    tibble(
      parameter = p,
      method    = paste0("Split-Half ", session_label),
      estimate  = spearman_brown(r),
      se        = se_r * 2 / (1 + r)^2, # delta method through the Spearman-Brown transform
      lower     = spearman_brown(ct$conf.int[1]),
      upper     = spearman_brown(ct$conf.int[2]),
      width     = spearman_brown(ct$conf.int[2]) - spearman_brown(ct$conf.int[1])
    )
  })
}

splithalf = bind_rows(
  splithalf_session(fit_bandit_t1_first, fit_bandit_t1_second, ids_splithalf, "(T1, N = 115)"),
  splithalf_session(fit_bandit_t2_first, fit_bandit_t2_second, ids_splithalf, "(T2, N = 115)")
)

splithalf |>
  gt() |>
  fmt_number(decimals = 3)
```

## Full T1 sample split-half (N = 547)

Same first-half/second-half split-half approach as above, but applied to the full N = 547 T1 sample instead of the N = 115 test-retest subsample -- there's no session 2 for these extra subjects, so this is the only split-half estimate available at this N. Run `02_01_pike_2026_fit_bandit_fullsample.R` first to produce these fits (it fits the split halves before the full-length fit used for RMU above).

```{r full-sample-splithalf}
splithalf_full_sample = splithalf_session(
  fit_bandit_fullsample_first, fit_bandit_fullsample_second, ids_all,
  paste0("(Full T1 sample, N = ", length(ids_all), ")")
)

splithalf_full_sample |>
  gt() |>
  fmt_number(decimals = 3)
```

# Plots Comparing Performance across estimators

## All estimates

```{r compare-reliability-plot}
fmt_no_leading_zero = function(x) sub("^(-?)0\\.", "\\1.", sprintf("%.2f", x))

comparison = bind_rows(rmu, testretest, splithalf, rel_full_sample, splithalf_full_sample) %>%
  arrange(parameter, method) |>
  mutate(
    parameter = case_when(
      parameter == "ivT" ~ "Inverse\n Temperature",
      parameter == "lrP" ~ "Punishment\n Learning Rate",
      parameter == "lrR" ~ "Reward\n Learning Rate"
    ),
    is_testretest = grepl("Test-Retest", method),
    method = fct_relevel(method, "Test-Retest (N = 115)"),
    label = paste0(
      fmt_no_leading_zero(estimate), " [",
      fmt_no_leading_zero(lower), ", ",
      fmt_no_leading_zero(upper), "]"
    )
  )

comparison %>%
  ggplot(aes(x = parameter, y = estimate, color = method, shape = method, size = is_testretest)) +
  geom_pointrange(aes(ymin = lower, ymax = upper), position = position_dodge(width = 0.8)) +
  geom_text(
    aes(y = lower, label = label, group = method),
    position = position_dodge(width = 0.8),
    angle = 90, hjust = 1.11, vjust = 0.5,
    size = 2, show.legend = FALSE
  ) +
  scale_shape_manual(values = c(20, 15, 17, 18, 12, 4, 8)) +
  scale_size_manual(values = c(`TRUE` = 1.3, `FALSE` = 0.7), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.25, 0.05))) +
  labs(
    x = NULL,
    y = "Reliability Estimate", color = "Method", shape = "Method",
    title = NULL,
    subtitle = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(colour = "black", size = 11),
    axis.title = element_text(size = 11)
    )

ggsave("02_pike_reliability_comparson_all.pdf", width = 6.5, height = 4)
```


## Filtering out T2 split-half/rmu estimates

```{r compare-reliability-plot-no-t2}
comparison_no_t2 = comparison %>%
  filter(!grepl("T2", method)) %>%
  mutate(method = fct_drop(method))

comparison_no_t2 %>%
  ggplot(aes(x = parameter, y = estimate, color = method, shape = method, size = is_testretest)) +
  geom_pointrange(aes(ymin = lower, ymax = upper), position = position_dodge(width = 0.8)) +
  geom_text(
    aes(y = lower, label = label, group = method),
    position = position_dodge(width = 0.8),
    angle = 90, hjust = 1.11, vjust = 0.5,
    size = 2, show.legend = FALSE
  ) +
  scale_shape_manual(values = c(8, 19, 1, 17, 24, 4, 8)) +
  scale_size_manual(values = c(`TRUE` = 1.3, `FALSE` = 0.7), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.25, 0.05))) +
  labs(
    x = NULL,
    y = "Reliability Estimate", color = "Method", shape = "Method",
    title = NULL,
    subtitle = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(colour = "black", size = 10),
    axis.title = element_text(size = 10)
    )

ggsave("02_pike_reliability_comparson_no_t2.pdf", width = 6.5, height = 4)
```

# Appendix: the Stan model

## Source

The stan model used to fit the Bayesian models is presented below:

```{r print-stan-model}
#| echo: false
#| results: asis

setwd("..") # set the working directory to the parent folder which has /data/

stan_path = file.path(data_dir, "model", "bandit_models", "model_3_noncentered.stan")
cat("``` stan\n", paste(readLines(stan_path), collapse = "\n"), "\n```\n", sep = "")
```

## How it works

Task: A hierarchical Rescorla-Wagner learner for a 4-armed bandit, where each arm can independently deliver a reward and a punishment on the same trial. 

Model: Every subject gets their own `lrR` (reward learning rate), `lrP` (punishment learning rate) and `ivT` (inverse-temperature / choice precision), and those are drawn from a population distribution estimated jointly with them -- so the group and the individuals are fit in one pass, and each subject is shrunk toward the group by an amount the data decides.


### `model` block -- the likelihood, one trial at a time

Model Structure:

1. **Expected rewards and punishments**

The model gives each subjec three reward value vectors of length 4 (value for each option), which are updated during th task. The model assumes that reward values start at `initV = [0,0,0,0]`:

   - `v_rwd` -- learned reward expectation of each arm
   - `v_plt` -- learned punishment expectation of each arm
   - `v` -- thier sum, which we'll call *net value*
2. **Prediction errors** 

- `peR = rwd[i,t] - v_rwd[choice]` actual reward value `rwd` (1/0) minus the reward expectation
- `peP = -abs(plt[i,t]) - v_plt[choice]` actual punishment value (-1/0) minus the punishment expectation. 

Note that `-abs()` forces the punishment target to be 0 or -1 regardless of how `plt` happens to be signed in the data file.

3. **Updating expectations**

- `v_rwd[choice[i,t]]= v_rwd[choice[i,t]] + lrR[i] * peR`
- `v_plt[choice[i,t]]= v_plt[choice[i,t]] + lrP[i] * peP`

Participants update their reward/punishment expectations using the prediction error (peR/peP) multiplied by the learning rates (lrR/lrP).

Only the expectations for the chosen option is updated - see note below. 

4. **Estimate net value**: 

- `v[choice] = v_rwd[choice] + v_plt[choice]`

The net value `v` is the sum of the reward and punishment expectations, and drives what subjects select (see next step).

5. **Selection Probability**: 

- `choice[i,t] ~ categorical_logit(ivT[i] * v)`. 

The probability of selecting each stimulus is modelled using a softmax link function. 

The argumment of the function is the net value (`v`) multiplied by the inverse temperature (`ivT`). When $ivT = 0$, participants will select each option randomly. As $ivT \to \infty$, participants will always choose the option with the higher reward expectation.

**Note.** The three unchosen arms keep whatever expected value they had. There is no fictive/counterfactual updating and no decay. This is the single biggest structural difference from hBayesDM's `bandit4arm_4par` (`02_03_pike_2026_bandit4arm_4par.qmd`), which decays *all four* arms toward zero every trial.

The loop is written per-subject with local `v_rwd`/`v_plt`/`v` declared inside the `for (i in 1:N)` body, so state resets at each subject automatically -- there's no cross-subject leakage even though one `model` block covers all 115 people.

### `data` block -- what a trial looks like

Everything is stored subject x trial, built by the authors' `bandit_datalist()` (`utils.R`) and shipped as `bandit_t1.json` / `bandit_t2.json`:

| Variable | Shape | Meaning |
|---|---|---|
| `N`, `T` | int | subjects, max trials |
| `Tsubj` | `N` | trials actually completed per subject; the trial loop runs `1:Tsubj[i]`, so shorter subjects just stop early and the trailing padding in the matrices is never read |
| `No` | int (4) | number of arms |
| `choice` | `N x T` | which arm (1-4) was chosen |
| `rwd` | `N x T` | reward on that trial, 1 or 0 |
| `plt` | `N x T` | punishment on that trial, -1 or 0 |
| `Vinits` | `No` | starting values, `[0, 0, 0, 0]` for reward expectations|

`Nopt` and `unchosen` are declared but never used here. They're inherited from the authors' fuller `model_5.stan`, which updates unchosen arms too; this model only ever touches the arm that was picked. Stan requires them in the data list anyway, since an unused declaration is still a declaration.

### `parameters` / `transformed parameters` -- the non-centered trick

The three parameters we actually care about are bounded: two learning rates on \[0, 1\] and an inverse-temperature `ivT` on \[0, 6\]. They aren't sampled on those scales. Instead each subject gets an unconstrained standard-normal score, and the population location/scale is applied afterwards:

```
lrR[i] = Phi_approx(mu[1] + sigma[1] * lrR_raw[i])
lrP[i] = Phi_approx(mu[2] + sigma[2] * lrP_raw[i])
ivT[i] = Phi_approx(mu[3] + sigma[3] * ivT_raw[i]) * 6
```

with `lrR_raw`, `lrP_raw`, `ivT_raw ~ std_normal()`. `Phi_approx()` is a fast approximation to the standard normal CDF, so it maps the real line onto (0, 1); multiplying by 6 stretches `ivT` onto (0, 6). `mu` and `sigma` are therefore the population mean and SD on the **probit scale**, not on the learning-rate scale -- `mu[1] = 0` means a population centred on `lrR = 0.5`, and to read `sigma` you have to push it through `Phi_approx` too.

I've tested using a multivariate normal prior for all three _raw variables, but this let to sampling issues. 

# Session Info 
  
```{r}

sessionInfo()

```
  
